The Markov Link Method: a nonparametric approach to combine observations from multiple experiments

### 2018-08-16 paper/paper.tex
they're not parameters anymore!

\newcommand{\plx}{\ensuremath{f}}
\newcommand{\pxy}{\ensuremath{g}}
\newcommand{\ply}{\ensuremath{h}}
% \newcommand{\DX}{\ensuremath{{\mathcal{D}^X}}}
% \newcommand{\DY}{\ensuremath{{\mathcal{D}^Y}}}
\newcommand{\DX}{\ensuremath{\mathcal{D}^X}}
\newcommand{\DY}{\ensuremath{\mathcal{D}^Y}}
\newcommand{\MLMH}{\ensuremath{\mathcal{A}}}

@@

\begin{abstract}
Using observations from multiple experiments is essential to building larger scientific theories.  For example, we might analyze different aspects of human neurons, such as morphology, transcriptomics, or electrophysiology.  A strong theory would account for each perspective as well as the relationship between the perspectives.  Unfortunately, we rarely have a single experiment that simultaneously measures every viewpoint, uses a sufficient sample size, and gets sufficient measurement precision.  In practice, we must therefore conduct different experiments and find a way to link the results.  The Markov Link Method (MLM) offers an algorithm to reason about these `links.' The method is nonparametric: it makes no assumptions about what the link should look like.  We evaluate the MLM on a pair of single-cell RNA techniques and gain new insight about the link between the two methods.
\end{abstract}

Imagine we have a collection of different experiments about human cells.  We would like to link the different perspectives that we get from each experiment.  To make this idea rigorous, we need to define what we mean by `perspective' and what we mean by `link.'

To define `perspective,' we assume each experiment has two aspects:
%
\begin{itemize}
    \item A \emph{selection procedure} defines how we select cells.  For example, one experiment might select only liver cells, and another might select only cells which express the Vip protein.  
    \item A \emph{measurement procedure} defines what we measure about each cell.  For example, one experiment might measure the transcriptomic profile of a cell, and another experiment might measure the cell morphology.
\end{itemize}
%
The selection procedure defines what the experiment studies, and the measurement procedures determines the experiment's `perspective.'

The `link' between two measurement procedures should enable us to answer questions such as ``I have observed the transcriptome of this cell -- what morphologies might this cell have?''  We formalize this idea through what we call the `measurement link parameters,' $\pxy(y|x)$.  Specifically, let us say we obtain measurement result $x$ from one measurement procedure.  The number $\pxy(y|x)$ indicates the probability of obtaining result $y$ from the second measurement procedure applied to measure the same specimen.

We here propose the The Markov Link Method (\MLM), an algorithm that uses a collection of experiments to estimate the measurement link parameters, \pxy.  In this paper we focus particularly on the case that we have experiments based on several selection procedures and exactly two different measurement procedures.  We also suppose that the measurement procedures return categorical observations, i.e.\ a `measurement' assigns a specimen to one of a finite set of categories.  The \MLM{} could also generalize to numerical observations, but in that case it would be appropriate to incorporate some knowledge of smoothness (e.g.\ through a Gaussian assumption).  Categorical data requires no additional modeling assumptions, so for simplicity we will here focus on the categorical case.  This setup, sketched in Figure \ref{fig:bigpicture}, is perhaps the simplest example that highlights all the important assumptions and properties of the \MLM.  It embodies a common situation in biological problems, such as is found in \cite{gouwens2018classification}.

@@

The Markov Link Method is a general way to estimate measurement link parameters from a collection of experiments.  For concreteness, we will use experiments about human cells as a running example.  Let us say that we have gathered many human cells.  For each cell we are interested in 
%
\begin{enumerate}
    \item $\ell$, the selection procedure used to obtain the specimen
    \item $X$, the cell's `transcriptomic type,' as measured through a transcriptomic analysis.  
    \item $Y$, the cell's `morphological type,' as measured by looking at an image of the cell.
\end{enumerate}
%
Our task is to link the transcriptomic and morphological perspectives on `cell-type,' as manifested through the $X$ and $Y$ measurement procedures.  For example, we might like to say ``This cell has transcriptomic type 3, therefore it probably has morphological type F.''  Our main assumption is that these kinds of associations are unaffected by the choice of selection procedure.  In particular:

@@

More formally, our task is to estimate the link parameters $\pxy$ from a collection of experiments.  We suppose that for each selection procedure we have two experiments: one that measured transcriptomics and another that measured morphology.  We have no experiments in which both types were measured for the same specimens.  This setup is sketched in Figure \ref{fig:bigpicture}.  Different setups could also be considered.   For example, we might additionally have a small experiment in which both measurement procedures could be applied to the same specimen.  We might have more than two measurement procedures.  We might have measurement procedures that yield numerical observations instead of categorical ones.  Generalizing the method to all these cases should be straightforward, but we leave it for future work.

In the specific case we will focus on, all of the experimental data can be summarized with two matrices, $\DX$ and $\DY$.  Here $\DX_{\ell x}$ indicates the number of specimens gathered with procedure $\ell$ that have transcriptomic type $x$.  Likewise $\DY_{\ell y}$ indicates the number of specimens gathered with procedure $\ell$ that have morphological type $y$.  For each transcriptomic type $x$ and morphological type $y$, the Markov Link Method uses the data to produce two objects:
%
\begin{enumerate}
    \item $\hat \pxy(y|x)$ -- a point estimate for the link parameter $\pxy(y|x)$
    \item $C_{x,y}$ -- a credible interval which contains the true link parameter $\pxy(y|x)$ with high probability
\end{enumerate}
%
To estimate these, we must consider certain `nuisance parameters.'  Our observable data, $\DX,\DY$, is governed by the following parameters:
\begin{itemize}
    \item $\plx(x|\ell)$ -- the probability that a cell sampled with procedure $\ell$ will have transcriptomic type $x$
    \item $\ply(y|\ell)$ -- the probability that a cell sampled with procedure $\ell$ will have morphological type $y$
\end{itemize}
It is straightforward to estimate $\plx,\ply$ from $\DX,\DY$ using standard methods.  We would like to use these parameters to tell us something about our object of interest, the link $\pxy$.  Under the Markov Link Method assumption, this link is connected to $\plx,\ply$ through the equations
\begin{gather}\label{eq:mlmh}
\sum_x \plx(x|\ell)\pxy(y|x)=\ply(y|\ell) \qquad \forall y,\ell
\end{gather}
With these equations and knowledge of $\plx,\ply$, we can produce an estimate for the link: $\hat \pxy_{\plx,\ply}$.  We can also produce bounds for the link.  We define these objects precisely in Appendix \ref{sec:mlmdetails}.  However, in practice we do not actually have exact knowledge of $\plx,\ply$.  We only have estimates from data.  We take a Bayesian perspective to account for our uncertainty in this estimation.  For our prior beliefs, we assume a noninformative uniform prior $\mathbb{P}$.  Following the Bayesian philosophy, we then incorporate new knowledge by conditioning.  We have two important pieces of knowledge about $\plx,\ply$.  First, we have observed the data, $\DX,\DY$.  Second, we know from the MLM assumption that there exists \emph{some} value $\pxy$ such that Equation (\ref{eq:mlmh}) holds; let \MLMH{} denote the event that this holds.  We then define our point estimate $\hat \pxy$ as the posterior expectation of $\hat \pxy_{\plx,\ply}$ after conditioning on this knowledge, i.e.
\[
\hat \pxy(y|x) = \mathbb{E}\left[\hat \pxy_{\plx,\ply} | \DX,\DY,\MLMH \right]
\]
We can use a similar strategy to produce the credible intervals $C_{x,y}$.  Exact details can be found in Appendix \ref{sec:mlmdetails}.  Code to compute $\hat \pxy(y|x)$ and $C_{x,y}$ is published at https://github.com/jacksonloper/markov-link-method, including a tutorial-style ipython notebook detailing every computation made in this paper.

@@

\caption{{\bf When can the MLM accurately estimate the link?} \label{fig:simI}  We use simulations to test the MLM.  In subfigure (a) we look at the error of the MLM link estimator in many different situations, varying the numbers of samples, selection procedures, and the link parameters.  In each trial our `ground truth' link parameters picked uniformly at random from the parameter space.  The MLM estimator accurately recovers the ground truth from simulated data, as long as we have enough samples \emph{and} enough distinct selection procedures.  In subfigure (b) we consider the MLM's credible intervals.  For a selection of trials we choose a random value of of $x,y$ and then compare the true link parameter $\pxy(y|x)$ with the MLM interval $C_{x,y}$.  We show the interval (in black) centered around the ground truth (in red).  Since the link parameters lie in $[0,1]$, the interval is always contained within $\pm1$ of the ground truth (this larger region is shown in gray).  The intervals mostly contain the truth, and get narrower with more samples as long as there are enough distinct selection procedures.   In subfigure (c) we focus on the case of exactly $4$ selection procedures.  In this case it is only possible to get very accurate estimates if the link takes on a particular form.  If the link is independent (i.e.\ $X,Y$ are independent) or if the link is invertible and deterministic (e.g.\ $X=Y$) then we can determine what the link is.  However, if the truth is more complex, small numbers of selection procedures may make it impossible to determine the true link; this issue is discussed in detail in Section \ref{sec:casestudies}.}
}

There are three desiderata we might hope the Markov Link Method estimators would achieve:
\begin{description}
    \item[Estimator convergence] As we obtain more samples, $\hat \pxy(y|x)$ should converge to the true link $\pxy(y|x)$ for each $x,y$.
    \item[Interval concentration] As we obtain more samples, the interval $C_{x,y}$ should get smaller. Asymptotically it should include nothing but the point $\pxy(y|x)$.
    \item[Conservative coverage] $\pxy(y|x) \in C_{x,y}$ with high probability.
\end{description}
However, due to certain identifiability issues, it is only actually possible to guarantee the last desideratum.  In many cases, the MLM achieves all three.  However, we can show that identifiability issues block any possibility of success in some cases; we detail this problem in Appendix \ref{sec:casestudies}.  The good news is that even though the Markov Link Method credible intervals may not concentrate around the truth, they do as well as possible.  In particular, they do appear to converge around the quantities we can estimate, namely the $q_\mathrm{lo},q_\mathrm{hi}$ bounds described in Appendix \ref{sec:mlmdetails}.  A precise statement of what we have been able to prove may be found in Theorem \ref{thm:mainthm} of Appendix \ref{sec:casestudies}.

The following two simulations give a good overview of the relevant issues:

\begin{itemize}
    \item How does MLM performance depend on the number of samples and the number of selection procedures?  We suppose we have two measurement procedures which return categorical measurements among six categories.  Thus the first procedure yields $X\in \{1,2,3,4,5,6\}$ and the second yields $Y\in \{1,2,3,4,5,6\}$.  To see how the method performs in different circumstances, we will run many trials.  In each trial we fix the number of selection procedures and pick `ground truth' parameters uniformly at random from the parameter space.  We then simulate a dataset from these ground truth parameters, fixing the total number of samples (spreading these samples equally among all combinations of selection procedure and measurement procedure).  Finally, we apply the MLM to the simulated data to get the point estimates $\tilde q(y|x)$ and the credible intervals $C_{x y}$.  We measure the overall estimator convergence using a kind of total variation distance:
    \[
    \mathrm{Error}(\tilde q,\pxy)=\frac{1}{12}\sum_{x=1}^{6}\sum_{y=1}^{6} |\tilde q(y|x) - \pxy(y|x)|
    \]
    This error ranges between zero and one.  Zero error indicates that $\tilde q=\pxy$. An error of one indicates that the estimate has completely incorrect beliefs about the probability mass, i.e. $\pxy(y|x)=0$ whenever $\tilde q(y|x)>0$ and vice-versa.  We also look at the MLM credible intervals and see how often they cover the true parameters.

    The results are summarized in Figure \ref{fig:simI}.  In trials with more samples, the estimator generally has lower error.  However, to make the error actually converge to zero we need at least six distinct selection procedures.  This figure also shows that the intervals work correctly regardless of of the number of samples or selection procedures; they include the ground truth with high probability.   With many samples and selection procedures, the intervals are small and concentrated around the truth.  With fewer samples or selection procedures, the intervals are sometimes forced to be larger.  However, our uncertainty is not the same for every parameter.  In some cases we are able to obtain a very tight credible interval for one particular parameter even though the overall estimator error is high.

    \item How does estimator convergence depend upon the link parameters themselves?  In each trial of this simulation we use four selection procedures and our measurement procedures yield one of six categories.  In the previous simulations we saw that estimator convergence was generally impossible in this situation, due to the small number of selection procedures.  However, in our previous simulations we picked the link parameters uniformly from parameter space.  Now we will be more choosey.   On one extreme, we will produce trials where the link makes $X,Y$ independent, i.e. $\pxy(y|x)=\pxy(y|x')$ for every $x,x'$.  On the other extreme, we will have trials where $X,Y$ are deterministically related by the equation $X=Y$.  We will also consider every link ``in-between'' these two extremes (found by convex combinations).  Figure \ref{fig:simI} shows that estimator convergence is actually possible in the two extreme cases.  However, estimator convergence fails for the in-between cases.  We leave a complete mathematical understanding of this result to future work.  For the purposes of this paper, we content ourselves by noting one interesting case: let the measurements return one of $2^k$ categories and let us use only $k+1$ carefully chosen selection procedures.  Now suppose the link defines any invertible deterministic function between $X$ and $Y$.  In this case, with enough enough samples, we can determine both that the relationship is deterministic and the exact specification of the invertible function.  This result is proven in Appendix \ref{sec:casestudies}, Theorem \ref{thm:miracle}.
\end{itemize}

@@

\figgy{
\hfill{}\figgygr{0.9}{../images/allenout}\hfill{}
\caption{{\bf MLM estimation of the link, with credible interval.}  The central plot shows a portion of the MLM estimator applied to the Allen Institute data.  We look at a subselection of Standard types ($x$) and all of the Patch types ($y$).  For each combination $x,y$ we draw a rectangle whose color indicates the value of the estimator $\hat \pxy(y|x)$.  We also show our confidence about these estimators.  On the left we indicate the lower bounds indicated by the credible interval, on the right we indicate upper bounds.  For some parameters our intervals are much tighter than others.  For example, it appears we have high confidence the Standard type `Vip Rspo4' is highly associated with the Patch type `Vip.'  By constrast, we have almost no idea what is associated with the Standard type `Meis2.'  The upper credible interval bounds suggest that $\pxy(y|\mathrm{Meis2})$ could be nearly 1 for many different Patch types.  Obviously it cannot be 1 for all of those types simultaneously, since $\sum_y \pxy(y|\mathrm{Meis2})=1$, but the data simply doesn't tell us which $y$ carries the mass.  \label{fig:allenboot}}
}

We applied the MLM to this data to obtain point estimates $\hat \pxy(y|x)$ and credible intervals $C_{x y}$.  In Figure \ref{fig:allenboot} we visualize these objects for selected values of $x,y$.  Each parameter has a slightly different story.  For example, for the Standard type $x=$`Vip Rspo4' and the Patch type $y=$`Vip,' we have that $C_{x,y}=[.88,1.0]$.  This supports the idea that the true link satisfies $\pxy(y|x)\geq .88$: at least 88\% of the cells classified as type `Vip Rspo4' by the Standard method will be classified as `Vip' by the Patch method.  However, for other types there is more ambiguity.  For the Standard type $x=$`Lamp5 Egln 1' and the Patch types $y=$`Lamp' we have $C_{x y}=[0,1.0]$.  The data we have does not give us a definitive answer as to whether cells with Standard type `Lamp5 Egln 1' are being classified with Patch type `Lamp.'

@@

In this paper we suggest one way to overcome these difficulties.  We formalize the `relationship' between two measurement procedures through the notion of a `measurement link.'  Given that we have measured a specimen with one procedure and obtained the outcome $x$, the link parameter $q(y|x)$ tells us the probability that we would obtain outcome $y$ if we measured the same specimen with the second procedure.  We present the Markov Link Method assumption, which roughly states that the link can be statistically isolated from the selection procedures.  We provide the Markov Link Method (MLM) algorithm that uses this assumption to estimate the link and measure our uncertainty.  Code is published at https://github.com/jacksonloper/markov-link-method, including a tutorial-style ipython notebook detailing every computation made in this paper.

@@

Another direction for future work would be to take a frequentist point of view.  An obvious direction would be to use profile likelihoods as statistics to define confidence intervals for the link parameters.  However, the distribution of these statistic is difficult to pin down.  Traditional techniques based on asymptotic normality may fail dramatically, because the true parameters may lie at the boundary of the parameter space.  In particular, there may be some $x,y$ for which $\pxy(y|x)=0$.  Even approximate methods are difficult to apply because the parameter space can be quite high-dimensional.  Perhaps future work could uncover a way to overcome these challenges.

@@

Consider experiments yielding an $\Omega_\ell\times\Omega_X$ matrix $\DX$ and an $\Omega_\ell \times \Omega_Y$ matrix $\DY$, carrying the distribution
\begin{align*}
(\DX_{\ell 1},\DX_{\ell 2}\cdots \DX_{\ell \Omega_X}) \sim \mathrm{Multinomial}(n_\ell,\plx(\cdot | \ell))\\
(\DY_{\ell 1},\DY_{\ell 2}\cdots \DY_{\ell \Omega_Y}) \sim \mathrm{Multinomial}(m_\ell,\ply(\cdot | \ell))
\end{align*}
where $\plx(x|\ell),\ply(y|\ell)$ are conditional distributions and there is some $\pxy(y|x)$ such that $\ply(y|\ell)=\sum_x \plx(x|\ell)\pxy(y|x)$.  This is simply a restatement of the assumptions we have made throughout this paper about how our parameters of interest $(\plx,\pxy,\ply)$ are related to the data we can observe $(\DX,\DY)$.

The purpose of the MLM is to make estimates about $\pxy$ using the data $\DX,\DY$.  Unfortunately, $\pxy$ cannot be directly determined from the data.  Even perfect knowledge of $\plx,\ply$ may be insufficient to determine the true value of $\pxy$.  Some examples are detailed in Appendix \ref{sec:casestudies}.  This problem is called `nonidentifiability,' and it can have some troubling consequences.  For example, standard Bayesian analyses applied to nonidentifiable parameters will be extremely sensitive to the precise choice of our prior beliefs.  Even with infinite data, the prior beliefs may have a significant impact on inferences.  To avoid these difficulties, we focus on parameters that we know we can identify from data.  In particular, we will look at lower bounds, upper bounds, and something in-between.

>>>>

\newcommand{\plx}{\ensuremath{f}}
\newcommand{\pxy}{\ensuremath{g}}
\newcommand{\ply}{\ensuremath{h}}
% \newcommand{\DX}{\ensuremath{{\mathcal{D}^X}}}
% \newcommand{\DY}{\ensuremath{{\mathcal{D}^Y}}}
\newcommand{\DX}{\ensuremath{{\mathcal{D}_X}}}
\newcommand{\DY}{\ensuremath{{\mathcal{D}_Y}}}
% \newcommand{\DX}{\ensuremath{\mathcal{D}^X}}
% \newcommand{\DY}{\ensuremath{\mathcal{D}^Y}}
\newcommand{\MLMH}{\ensuremath{\mathcal{A}}}

@@

\begin{abstract}
We propose the Markov Link Method (\MLM), a way to synthesize experiments based on several measurement procedures.  For example, we might have different experiments measuring different aspects of human neurons, such as morphology, transcriptomics, or electrophysiology.  The \MLM{} uses this data to link these different aspects.  It answers the question ``We have observed the transcriptome of this cell -- what might its morphology have been?'' The method is nonparametric: it makes no assumptions about what the link looks like.  We evaluate the MLM on a pair of single-cell RNA techniques and gain new insight about the link between the two methods.
\end{abstract}

We here propose the Markov Link Method (\MLM), a technique to understand how different measurement procedures are related.  For our purposes, we assume each experiment has two aspects:
%
\begin{itemize}
    \item A \emph{selection procedure} defines how we select cells.  For example, one experiment might select only liver cells, and another might select only cells which express the Vip protein.  
    \item A \emph{measurement procedure} defines what we measure about each cell.  For example, one experiment might measure the transcriptomic profile of a cell, and another experiment might measure the cell morphology.
\end{itemize}
%
To synthesize experiments with different measurement procedures, we must understand how the measurement procedures are related.  For example, we would like to be able to answer questions such as ``We have measured the transcriptome of this cell -- what might its morphology have been?''  We formalize this idea through what we call the `measurement link,' $\pxy(y|x)$.  Specifically, let us say we obtain measurement result $x$ from one measurement procedure.  The number $\pxy(y|x)$ indicates the probability of obtaining result $y$ from the second measurement procedure applied to measure the same specimen.

We here propose the The Markov Link Method (\MLM), an algorithm that uses a collection of experiments to estimate the measurement link, \pxy.  In this paper we focus particularly on the case that we have experiments based on several selection procedures and exactly two different measurement procedures.  We also suppose that the measurement procedures return categorical observations, i.e.\ a `measurement' assigns a specimen to one of a finite set of categories.  The \MLM{} could also generalize to numerical observations, but in that case it would be appropriate to incorporate some knowledge of smoothness (e.g.\ through a Gaussian assumption).  Categorical data requires no additional modeling assumptions, so for simplicity we will here focus on the categorical case.  This setup, sketched in Figure \ref{fig:bigpicture}, is perhaps the simplest example that highlights all the important assumptions and properties of the \MLM.  It embodies a common situation in biological problems, such as is found in \cite{gouwens2018classification}.

@@

The Markov Link Method is a general way to estimate measurement link from a collection of experiments.  For concreteness, we will use experiments about human cells as a running example.  Let us say that we have gathered many human cells.  For each cell we are interested in 
%
\begin{enumerate}
    \item $\ell$, the selection procedure used to obtain the specimen
    \item $X$, the cell's `transcriptomic type,' as measured through a transcriptomic analysis.  
    \item $Y$, the cell's `morphological type,' as measured by looking at an image of the cell.
\end{enumerate}
%
Our task is to link the transcriptomic and morphological perspectives on `cell-type,' as manifested through the $X$ and $Y$ measurement procedures.  For example, we might like to say ``This cell has transcriptomic type 3, therefore it probably has morphological type F.''  Our main assumption is that these kinds of associations are unaffected by the choice of selection procedure.  In particular:

@@

More formally, our task is to estimate the link $\pxy$ from a collection of experiments.  We suppose that for each selection procedure we have two experiments: one that measured transcriptomics and another that measured morphology.  We have no experiments in which both types were measured for the same specimens.  This setup is sketched in Figure \ref{fig:bigpicture}.  Different setups could also be considered.   For example, we might additionally have a small experiment in which both measurement procedures could be applied to the same specimen.  We might have more than two measurement procedures.  We might have measurement procedures that yield numerical observations instead of categorical ones.  Generalizing the method to all these cases should be straightforward, but we leave it for future work.

In the specific case we will focus on, all of the experimental data can be summarized with two matrices, $\DX$ and $\DY$.  The matrix $\DX$ tabulates the transcriptomic types found in specimens gathered with each selection procedure; $\DY$ does the same for morphological types.   For each transcriptomic type $x$ and morphological type $y$, the Markov Link Method uses this data to produce two objects:
%
\begin{enumerate}
    \item $\hat \pxy(y|x)$ -- a point estimate for the link $\pxy(y|x)$
    \item $C_{x,y}$ -- a credible interval which contains the true link $\pxy(y|x)$ with high probability
\end{enumerate}
%
To estimate these, we must consider certain `nuisance' quantities.  The link is our primary target of study, but we cannot estimate the link without considering these other objects.  Our observable data, $\DX,\DY$, is governed by the following conditional distributions:
\begin{itemize}
    \item $\plx(x|\ell)$ -- the probability that a cell sampled with procedure $\ell$ will have transcriptomic type $x$
    \item $\ply(y|\ell)$ -- the probability that a cell sampled with procedure $\ell$ will have morphological type $y$
\end{itemize}
It is straightforward to estimate $\plx,\ply$ from $\DX,\DY$ using standard methods.  We would like to use these distributions to tell us something about our object of interest, the link $\pxy$.  Under the Markov Link Method assumption, this link is connected to $\plx,\ply$ through the equations
\begin{gather}\label{eq:mlmh}
\sum_x \plx(x|\ell)\pxy(y|x)=\ply(y|\ell) \qquad \forall y,\ell
\end{gather}
With these equations and knowledge of $\plx,\ply$, we can produce an estimate for the link: $\hat \pxy_{\plx,\ply}$.  If the equations are underdetermined, we can also produce bounds for the link.  We define these objects precisely in Appendix \ref{sec:mlmdetails}.  However, in practice we do not actually have exact knowledge of $\plx,\ply$.  We only have estimates from data.  We take a Bayesian perspective to account for our uncertainty in this estimation.  For our prior beliefs, we assume a noninformative uniform prior $\mathbb{P}$.  Following the Bayesian philosophy, we then incorporate new knowledge by conditioning.  We have two important pieces of knowledge about $\plx,\ply$.  First, we have observed the data, $\DX,\DY$.  Second, we know from the MLM assumption that there exists \emph{some} value $\pxy$ such that Equation (\ref{eq:mlmh}) holds; let \MLMH{} denote the event that this holds.  By conditioning on our knowledge, we obtain the posterior distribution $\mathbb{P}(d\plx,d\ply | \DX,\DY,\MLMH)$.  Our point estimate $\hat \pxy$ is then defined as a Bayes estimator for $\hat \pxy_{\plx,\ply}$ with respect to this posterior:
\[
\hat \pxy(y|x) \triangleq \mathbb{E}\left[\hat \pxy_{\plx,\ply} | \DX,\DY,\MLMH \right]
               = \int \hat \pxy_{\plx,\ply} \mathbb{P}(d\plx,d\ply | \DX,\DY,\MLMH) 
\]
This estimate minimizes the expected squared risk error under the posterior distribution.  We can use similar strategies to produce the credible intervals $C_{x,y}$.  Exact details can be found in Appendix \ref{sec:mlmdetails}.  Code to compute $\hat \pxy(y|x)$ and $C_{x,y}$ is published at https://github.com/jacksonloper/markov-link-method, including a tutorial-style ipython notebook detailing every computation made in this paper.

@@

\caption{{\bf When can the MLM accurately estimate the link?} \label{fig:simI}  We use simulations to test the MLM.  In subfigure (a) we look at the error of the MLM link estimator in many different situations, varying the numbers of samples, selection procedures, and the link itself.  In each trial our `ground truth' link picked uniformly at random from the parameter space.  The MLM estimator accurately recovers the ground truth from simulated data, as long as we have enough samples \emph{and} enough distinct selection procedures.  In subfigure (b) we consider the MLM's credible intervals.  For a selection of trials we choose a random value of of $x,y$ and then compare the true link $\pxy(y|x)$ with the MLM interval $C_{x,y}$.  We show the interval (in black) centered around the ground truth (in red).  Since the link probabilities lie in $[0,1]$, the interval is always contained within $\pm1$ of the ground truth (this larger region is shown in gray).  The intervals mostly contain the truth, and get narrower with more samples as long as there are enough distinct selection procedures.   In subfigure (c) we focus on the case of exactly $4$ selection procedures.  In this case it is only possible to get very accurate estimates if the link takes on a particular form.  If the link is independent (i.e.\ $X,Y$ are independent) or if the link is invertible and deterministic (e.g.\ $X=Y$) then we can determine what the link is.  However, if the truth is more complex, small numbers of selection procedures may make it impossible to determine the true link; this issue is discussed in detail in Section \ref{sec:casestudies}.}
}

There are three desiderata we might hope the Markov Link Method estimators would achieve:
\begin{description}
    \item[Estimator convergence] As we obtain more samples, $\hat \pxy(y|x)$ should converge to the true link $\pxy(y|x)$ for each $x,y$.
    \item[Interval concentration] As we obtain more samples, the interval $C_{x,y}$ should get smaller. Asymptotically it should include nothing but the point $\pxy(y|x)$.
    \item[Conservative coverage] $\pxy(y|x) \in C_{x,y}$ with high probability.
\end{description}
In many cases, the MLM achieves all three of these conditions.  In other cases, we can show that an issue known as `identifiability' blocks even the possibility of estimator convergence or interval concentration.  However, even when interval concentration is impossible, the \MLM{} appears to do as well as it possibly can: the intervals $C_{x,y}$ close around the bounds of what can possibly be learned from the data we observe.  We have begun to develop a theory of the relevant issues, which we detail in Appendix \ref{sec:mlmdetails}.  We give a few of the interesting corner-cases and prove a few surprising results.  A complete story remains elusive.  The following two simulations give a good overview of the relevant issues:

\begin{itemize}
    \item How does MLM performance depend on the number of samples and the number of selection procedures?  We suppose we have two measurement procedures which return categorical measurements among six categories.  Thus the first procedure yields $X\in \{1,2,3,4,5,6\}$ and the second yields $Y\in \{1,2,3,4,5,6\}$.  To see how the method performs in different circumstances, we will run many trials.  In each trial we fix the number of selection procedures and pick a `ground truth' link uniformly at random from the parameter space.  We then simulate a dataset from this ground truth link, fixing the total number of samples (spreading these samples equally among all combinations of selection procedure and measurement procedure).  Finally, we apply the MLM to the simulated data to get the point estimates $\tilde q(y|x)$ and the credible intervals $C_{x y}$.  We measure the overall estimator convergence using a kind of total variation distance:
    \[
    \mathrm{Error}(\tilde q,\pxy)=\frac{1}{12}\sum_{x=1}^{6}\sum_{y=1}^{6} |\tilde q(y|x) - \pxy(y|x)|
    \]
    This error ranges between zero and one.  Zero error indicates that $\tilde q=\pxy$. An error of one indicates that the estimate has completely incorrect beliefs about the probability mass, i.e. $\pxy(y|x)=0$ whenever $\tilde q(y|x)>0$ and vice-versa.  We also look at the MLM credible intervals and see how often they cover the true parameters.

    The results are summarized in Figure \ref{fig:simI}.  In trials with more samples, the estimator generally has lower error.  However, to make the error actually converge to zero we need at least six distinct selection procedures.  This figure also shows that the intervals work correctly regardless of of the number of samples or selection procedures; they include the ground truth with high probability.   With many samples and selection procedures, the intervals are small and concentrated around the truth.  With fewer samples or selection procedures, the intervals are sometimes forced to be larger.  However, our uncertainty is not the same for every aspect of the link.  In some cases we are able to obtain a very tight credible interval for one particular value even though the overall estimator error is high.

    \item How does estimator convergence depend upon the link itself?  In each trial of this simulation we use four selection procedures and our measurement procedures yield one of six categories.  In the previous simulations we saw that estimator convergence was generally impossible in this situation, due to the small number of selection procedures.  However, in our previous simulations we picked the link uniformly from its parameter space.  Now we will be more choosey.   On one extreme, we will produce trials where the link makes $X,Y$ independent, i.e. $\pxy(y|x)=\pxy(y|x')$ for every $x,x'$.  On the other extreme, we will have trials where $X,Y$ are deterministically related by the equation $X=Y$.  We will also consider every link ``in-between'' these two extremes (found by convex combinations).  Figure \ref{fig:simI} shows that estimator convergence is actually possible in the two extreme cases.  However, estimator convergence fails for the in-between cases.  We leave a complete mathematical understanding of this result to future work.  For the purposes of this paper, we content ourselves by noting one interesting case: let the measurements return one of $2^k$ categories and let us use only $k+1$ carefully chosen selection procedures.  Now suppose the link defines any invertible deterministic function between $X$ and $Y$.  In this case, with enough enough samples, we can determine both that the relationship is deterministic and the exact specification of the invertible function.  This result is proven in Appendix \ref{sec:casestudies}, Theorem \ref{thm:miracle}.
\end{itemize}

@@

\figgy{
\hfill{}\figgygr{0.9}{../images/allenout}\hfill{}
\caption{{\bf MLM estimation of the link, with credible interval.}  The central plot shows a portion of the MLM estimator applied to the Allen Institute data.  We look at a subselection of Standard types ($x$) and all of the Patch types ($y$).  For each combination $x,y$ we draw a rectangle whose color indicates the value of the estimator $\hat \pxy(y|x)$.  We also show our confidence about these estimators.  On the left we indicate the lower bounds indicated by the credible interval, on the right we indicate upper bounds.  For some aspects of the link our intervals are much tighter than others.  For example, it appears we have high confidence the Standard type `Vip Rspo4' is highly associated with the Patch type `Vip.'  By constrast, we have almost no idea what is associated with the Standard type `Meis2.'  The upper credible interval bounds suggest that $\pxy(y|\mathrm{Meis2})$ could be nearly 1 for many different Patch types.  Obviously it cannot be 1 for all of those types simultaneously, since $\sum_y \pxy(y|\mathrm{Meis2})=1$, but the data simply doesn't tell us which $y$ carries the mass.  \label{fig:allenboot}}
}

We applied the MLM to this data to obtain point estimates $\hat \pxy(y|x)$ and credible intervals $C_{x y}$.  In Figure \ref{fig:allenboot} we visualize these objects for selected values of $x,y$.  Each part of the link has a slightly different story.  For example, for the Standard type $x=$`Vip Rspo4' and the Patch type $y=$`Vip,' we have that $C_{x,y}=[.88,1.0]$.  This supports the idea that the true link satisfies $\pxy(y|x)\geq .88$: at least 88\% of the cells classified as type `Vip Rspo4' by the Standard method will be classified as `Vip' by the Patch method.  However, for other types there is more ambiguity.  For the Standard type $x=$`Lamp5 Egln 1' and the Patch types $y=$`Lamp' we have $C_{x y}=[0,1.0]$.  The data we have does not give us a definitive answer as to whether cells with Standard type `Lamp5 Egln 1' are being classified with Patch type `Lamp.'

@@

In this paper we suggest one way to overcome these difficulties.  We formalize the `relationship' between two measurement procedures through the notion of a `measurement link.'  Given that we have measured a specimen with one procedure and obtained the outcome $x$, the link $q(y|x)$ tells us the probability that we would obtain outcome $y$ if we measured the same specimen with the second procedure.  We present the Markov Link Method assumption, which roughly states that the link can be statistically isolated from the selection procedures.  We provide the Markov Link Method (MLM) algorithm that uses this assumption to estimate the link and measure our uncertainty.  Code is published at https://github.com/jacksonloper/markov-link-method, including a tutorial-style ipython notebook detailing every computation made in this paper.

@@

Another direction for future work would be to take a frequentist point of view.  An obvious direction would be to use profile likelihoods as statistics to define confidence intervals for the link.  However, the distribution of these statistic is difficult to pin down.  Traditional techniques based on asymptotic normality may fail dramatically, because the true link may lie at the boundary of the parameter space.  In particular, there may be some $x,y$ for which $\pxy(y|x)=0$.  Even approximate methods are difficult to apply because the parameter space can be quite high-dimensional.  Perhaps future work could uncover a way to overcome these challenges.

@@

Consider experiments yielding an $\Omega_\ell\times\Omega_X$ matrix $\DX$ and an $\Omega_\ell \times \Omega_Y$ matrix $\DY$, carrying the distribution
\begin{align*}
(\DX_{\ell 1},\DX_{\ell 2}\cdots \DX_{\ell \Omega_X}) \sim \mathrm{Multinomial}(n_\ell,\plx(\cdot | \ell))\\
(\DY_{\ell 1},\DY_{\ell 2}\cdots \DY_{\ell \Omega_Y}) \sim \mathrm{Multinomial}(m_\ell,\ply(\cdot | \ell))
\end{align*}
where $\plx(x|\ell),\ply(y|\ell)$ are conditional distributions and there is some $\pxy(y|x)$ such that $\ply(y|\ell)=\sum_x \plx(x|\ell)\pxy(y|x)$.  This is simply a restatement of the assumptions we have made throughout this paper about how our objects of interest $(\plx,\pxy,\ply)$ are related to the data we can observe $(\DX,\DY)$.

The purpose of the MLM is to make estimates about $\pxy$ using the data $\DX,\DY$.  Unfortunately, $\pxy$ cannot be directly determined from the data.  Even perfect knowledge of $\plx,\ply$ may be insufficient to determine the true value of $\pxy$.  Some examples are detailed in Appendix \ref{sec:casestudies}.  This problem is called `nonidentifiability,' and it can have some troubling consequences.  For example, standard Bayesian analyses applied to nonidentifiable parameters will be extremely sensitive to the precise choice of our prior beliefs.  Even with infinite data, the prior beliefs may have a significant impact on inferences.  To avoid these difficulties, we focus on objects that we know we can identify from data.  In particular, we will look at lower bounds, upper bounds, and something in-between.
